---
title: "DS 362 Homework 2"
format: 
  html:
    theme: pulse  
    toc: true
---

This is Homework Assignment 2 for DS 362. You can view the source code for this assignment on GitHub: [view the source code](https://github.com/jmgraham30/ds_362_assignments/blob/main/homework_02/index.qmd).

For your amusement: Knock knock. "Who's there?" Figs. "Figs who?" "Figs the doorbell. I've been knocking forever!"

## Problems

Many of the problems in this homework assignment are from the course text [An Introduction to Statistical Learning](https://www.statlearning.com/) which we will abbreviate as ISLR. [Access the book online here](https://hastie.su.domains/ISLR2/ISLRv2_corrected_June_2023.pdf).

Please complete and hand in this assignment as a Quarto notebook. 

### Problem 1

Please complete exercise 1 from section 2.4 on page 52 from the ISLR course text.


### Problem 2


Please complete exercise 4 from section 2.4 on page 53 from the ISLR course text.

### Problem 3


Please complete exercise 10 from section 2.4 on page 57 from the ISLR course text.


### Problem 4

Complete exercise 1 on pages 39 - 40 of the textbook Statistical Learning with Math and R by Joe Suzuki, available online through the Weinberg Library, [view the text](https://link-springer-com.ezp.scranton.edu/book/10.1007/978-981-15-7568-6).

This problem is walks you through the steps in the derivation of the linear least squares problems for simple linear regression. 


### Problem 5

Suppose that $\| \cdot \|$ is a vector norm and let ${\bf u}$ and ${\bf v}$ be vectors. Show that 

$$
| \|u\| - \|v\| | \leq \| u - v \|
$$

**Hint:** This can be done by noting that $\|{\bf u}\| = \|{\bf u} - {\bf v} + {\bf v} \|$ and then manipulating the triangle inequality. 

### Problem 6

Use the `qr` command in R to compute the $QR$ factorization of each of the following matrices. Write down the sizes of the $Q$ and $R$ matrices in each case.

a. $A = \left[\begin{array}{ccc} -1 & 4 & -3 \\ -1 & 0 & 1 \\ -6 & 2 & 3  \end{array}\right]$. Here is the R code to create this matrix:

```{r}
(A <- matrix(c(-1,4,-3,-1,0,1,-6,2,3),3,3,byrow = TRUE))
```

b. $B = \left[\begin{array}{ccc} -1 & 0 & 1 \\ -6 & 2 & 3  \end{array}\right]$. Here is the R code to create this matrix:

```{r}
(B <- matrix(c(-1,0,1,-6,2,3),2,3,byrow = TRUE))
```


c. $C = \left[\begin{array}{cc} -1  & -3 \\ -1 & 0 \\  2 & 3  \end{array}\right]$. Here is the R code to create this matrix:

```{r}
(C <- matrix(c(-1,-3,-1,0,2,3),3,2,byrow = TRUE))
```


