---
title: "DS 362 Modeling Case Study 2"
format: 
  html:
    theme: pulse
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(tidymodels)

theme_set(theme_minimal(base_size=13))

tidymodels_prefer()
```

For your amusement: Where do hamburgers like to dance? At a meatball! 

## Assignment Overview and Instructions

The goal of this modeling case study is for you to get hands-on experience in going through the steps of a machine learning problem from beginning to end with provided data. The case studies in this assignment will focus on hyperparameter tuning, fitting and comparing multiple models, and unsupervised methods that have been covered in class so far or are similar to those we have covered in class so far. 

For this assignment, you should create and work in an RStudio project named `case_study_2_lastname` where you will replace the `lastname` part with your last name. You should create either an appropriately names Quarto notebook or an appropriately names R script to save all you code in. Finally, you should maintain your project as a GitHub repository.


### Hyperparameter Tuning and Model Comparison Case Study

In this part of the assignment, you are asked to model a data set on sightings of [unidentified flying objects](https://en.wikipedia.org/wiki/Unidentified_flying_object) (UFOs). This data is modified from a [TidyTuesday](https://github.com/rfordatascience/tidytuesday/tree/master) data set available through [this repository](https://github.com/rfordatascience/tidytuesday/blob/master/data/2023/2023-06-20/readme.md) but originates from [the National UFO Reporting Center](https://nuforc.org/ndx/?id=shape).

To read in the data, run the following code:

```{r}
#| message: false
ufo_df <- read_csv("https://raw.githubusercontent.com/jmgraham30/modeling_case_study_2_data/main/ufo_df.csv")
```

Now, we can see what is contained in our data set:

```{r}
glimpse(ufo_df)
```

The data set contains the following variables:

- Response variable: `shape` - the shape of the UFO. This is a binary categorical variable that records whether the object for which the sighting was reported had a fixed geometric shape or some kind of amorphous dynamic shape.

- Predictor variables: 

    - `duration_seconds` - the duration of the sighting in seconds.
    
    - `day_part` - whether the reported sighting occurred in the morning, afternoon, or night. 
    
    - `latitude` - the latitude of the sighting.
    
    - `longitude` - the longitude of the sighting.
    
    - `population` - the population of the city in which the reported sighting occurred. 
    
    - `elevation_m` - the elevation in meters for the city in which the reported sighting occurred. 
    


### Unsupervised Learning Case Study


