---
title: "DS 362 Homework 3"
format: 
  html:
    theme: pulse  
    toc: true
---

```{r}
#| include: false
#| message: false
#| warning: false

library(tidyverse)
library(tidymodels)

theme_set(theme_minimal(base_size = 12))

```

This is Homework Assignment 3 for DS 362. You can view the source code for this assignment on GitHub: [view the source code](https://github.com/jmgraham30/ds_362_assignments/blob/main/homework_03/index.qmd).

For your amusement: What's orange and sounds like a parrot? A carrot!

## Problems

Many of the problems in this homework assignment are from the course text [An Introduction to Statistical Learning](https://www.statlearning.com/) which we will abbreviate as ISLR. [Access the book online here](https://hastie.su.domains/ISLR2/ISLRv2_corrected_June_2023.pdf).

Please complete and turn in this assignment as a Quarto notebook. 

### Problem 1

Complete exercise 3 on pages 121-122 from [An Introduction to Statistical Learning](https://www.statlearning.com/).

### Problem 2

Complete exercise 4 on page 122 from [An Introduction to Statistical Learning](https://www.statlearning.com/). Note that in that problem RSS is what we referred to as the squared error of residuals in lecture. 

### Problem 3

Complete exercise 8 on page 123 from [An Introduction to Statistical Learning](https://www.statlearning.com/). To do this exercise, you will need to install and load the `ISLR2` package in order to access the `Auto` data set. 

For now, you may omit (iv) from part (a). 

For part (b) rather than use `abline` to add the least squares regression line, use `geom_smooth` from the `ggplot2` package. Here is an example of how it works:

```{r}
#| echo: true
#| code-fold: false
#| message: false
#| warning: false

diamonds %>% 
  filter(carat <= 1) %>% 
  ggplot(aes(x=carat,y=sqrt(price))) + 
  geom_point() + 
  geom_smooth(method="lm",se=FALSE)
```

The important point is that `geom_smooth(method="lm",se=FALSE)` adds the least squares regression line which is specified by `method="lm"`, the `se=FALSE` argument removes the error band around the regression line which `geom_smooth` includes by default. 

### Problem 4

Complete exercise 9 on pages 123 - 124 from [An Introduction to Statistical Learning](https://www.statlearning.com/). To do this exercise, you will need to install and load the `ISLR2` package in order to access the `Auto` data set. 

For part (a), you can accomplish this using the `ggpairs` function from the `GGAlly` package which you will need to install and load. 

For now, you can ignore (ii) for part (c).